The state learner -- a super learner for right-censored data

### 2024-05-14 tex/jrss-version/state-learner-jrss-style.tex
uk spelling

\documentclass{statsoc}
\usepackage[a4paper]{geometry}

\usepackage{natbib, dsfont, enumitem,
  amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}

\usepackage[tablesonly,heads,nolists]{endfloat}

%% Theorem environment
% \theoremstyle{plain} % plain italic style
\newtheorem{theorem}{Theorem}
% \numberwithin{theorem}{section}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}

@@

\bibliographystyle{rss}

@@

%% TODO: Update figures with name 'figure-'.
%% Write macro for including figure or not.

% \newcommand{\includeFigCond}[2]{
%   \ifx\nofig\undefined %
%     \includegraphics[#1]{#2} %
%   \else %
%     \texttt{#2} %
%   \fi %
% }
\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

@@

\title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
\author{Anders Munch}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
\email{a.munch@sund.ku.dk}
\author[Munch {\it et al.}]{Thomas A.\ Gerds}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone
  tools and in applications of causal inference to estimate nuisance
  parameters. The super learner is a machine learning algorithm which
  combines a library of prediction models into a meta learner based on
  cross-validated loss. In right censored data, loss needs to be estimated somehow.
  % based on a model for the censoring distribution.
  % Unfortunately, the
  % commonly used partial likelihood loss is not suited for super learning, and
  % inverse probability of censoring weighted loss functions require a
  % pre-specified estimator of the censoring distribution. To relax this,
  We introduce the state learner, a new super learner for survival analysis, which
  simultaneously evaluates libraries
  of prediction models for the event of interest and the censoring
  distribution. The state learner can
be applied to all types of survival models, works in the presence
of competing risks, and does not require a single pre-specified
estimator of the conditional censoring distribution.
  We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
\end{abstract}

@@

@@

We consider the right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
that the event times and the censoring times are conditionally
independent given covariates, \( T \independent C \mid X \). This
implies that any distribution \( P \in \mathcal{P} \) is characterised
by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
function for censoring. For ease of presentation we now also assume
that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. The
distribution is characterized by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have \(H\)-almost everywhere
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}

@@

$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a distribution

@@

    \intertext{where in correspondance with  the relations in equation \eqref{eq:transition},} 

@@

\begin{corollary}
  \label{cor:asymp-cons}
  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \)
  and that there exists a sequence \( \phi_n \in \mathcal{F}_n \),
  \( n \in \N \), such that
  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
  \begin{enumerate}[label=(\roman*), topsep=0pt]
  \item If $\alpha=1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
      \right]} = \bigO(\log(n)n^{-1}) \).
  \item If $\alpha<1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
    \bigO(n^{-\alpha}) \).
  \end{enumerate}
\end{corollary}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figure~\ref{fig:zelefski}. We see that for
most sample sizes, the state learner selected prediction models for
both censoring and outcome which have similar or higher IPA compared
to the prediction models selected by \texttt{survSL}.
% ANDERS: Divide Fig 3 into two figures. keep y-axis fix for Outcome model panels but use different y-axes for  Censoring model.
\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
  \caption[]{For the risk prediction models of both the outcome and the
    censoring probability provided by each of the super learners, the IPA at the fixed time horizon is
    plotted against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  % ANDERS: add marginal censoring probability to panel titles
  \label{fig:zelefski}
\end{figure}

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{zelefski-real-data.pdf}
  \caption[]{The results of applying the 125 combinations of learners to the
    prostate cancer data set. The learners are \texttt{KM} (KM), \texttt{Cox
      strata CS} (strata), \texttt{Lasso} (lasso), \texttt{Elastic} (elastic),
    and \texttt{RF} (RF) as described Table~\ref{tab:zel-library}. The error
    bars are based on five repetitions using different splits. We refer to
    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
    learner', `Death learner', and `Censoring learner', respectively.}
  \label{fig:zelefski-real}
\end{figure}

@@

\begin{figure}
  \centering%
  \includeFigCond[width=1\linewidth]{zelefsky-data-target-par.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

@@

\appendix

@@

\bibliography{bib.bib}

>>>>

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
% \documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
% \documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
% \documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

\onecolumn % for one column layouts

%\usepackage{showframe}

% \graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
% \newtheorem{proposition}[theorem]{Proposition}%
% to get separate numbers for theorem and proposition etc.:
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

%% OLD

% \usepackage{natbib, dsfont,
% amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}

\usepackage{dsfont, booktabs}

% \usepackage[tablesonly,heads,nolists]{endfloat}

@@

% \bibliographystyle{rss}

@@

\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

@@

@@

%% NEW

\journaltitle{Journal of the Royal Statistical Society Series B: Statistical
  Methodology} %
\DOI{DOI HERE} %
\copyrightyear{xxxx} %
\pubyear{yyyy}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[The state learner]{The state learner -- a super learner for
  right-censored data}

\author[1,$\ast$]{Anders Munch}
\author[1]{Thomas A.\ Gerds}

\authormark{Munch et al.}

\address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
    Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}

\corresp[$\ast$]{Corresponding author. \href{email:email-id.com}{a.munch@sund.ku.dk}}

% \received{Date}{0}{Year}
% \revised{Date}{0}{Year}
% \accepted{Date}{0}{Year}

\abstract{In survival analysis, prediction models are needed as stand-alone
  tools and in applications of causal inference to estimate nuisance parameters.
  The super learner is a machine learning algorithm which combines a library of
  prediction models into a meta learner based on cross-validated loss. In right
  censored data, loss needs to be estimated somehow. We introduce the state
  learner, a new super learner for survival analysis, which simultaneously
  evaluates libraries of prediction models for the event of interest and the
  censoring distribution. The state learner can be applied to all types of
  survival models, works in the presence of competing risks, and does not
  require a single pre-specified estimator of the conditional censoring
  distribution. We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.}

@@

\maketitle

@@

We consider the right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
that the event times and the censoring times are conditionally
independent given covariates, \( T \independent C \mid X \). This
implies that any distribution \( P \in \mathcal{P} \) is characterised
by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
function for censoring. For ease of presentation we now also assume
that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. The
distribution is characterised by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have \(H\)-almost everywhere
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

  \includeFigCond[width=.3\textwidth]{figure-multi-state-process.pdf}

@@

$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a distribution

@@

    \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 

@@

\begin{corollary}
  \label{cor:asymp-cons}
  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \)
  and that there exists a sequence \( \phi_n \in \mathcal{F}_n \),
  \( n \in \N \), such that
  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
  \begin{enumerate}
  \item If $\alpha=1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
      \right]} = \bigO(\log(n)n^{-1}) \).
  \item If $\alpha<1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
    \bigO(n^{-\alpha}) \).
  \end{enumerate}
\end{corollary}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

@@

\begin{figure}
  \centering %
  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figures~\ref{fig:zelefski-out}
and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
learner selected prediction models for both censoring and outcome which have
similar or higher IPA compared to the prediction models selected by
\texttt{survSL}.
\begin{figure}
  \centering %
  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-survSL-out.pdf}
  \caption[]{For the risk prediction models of the outcome provided by each
    of the super learners, the IPA at the fixed time horizon is plotted against
    sample size. The results are averages across 1000 repetitions and the error
    bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-out}
\end{figure}

\begin{figure}
  \centering %
  \includeFigCond[width=.5\linewidth]{experiment-fig-sl-survSL-cens.pdf}
  \caption[]{For the risk prediction models of the censoring model provided by
    each of the super learners, the IPA at the fixed time horizon is plotted
    against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-cens}
\end{figure}

@@

\begin{figure}
  \centering %
  \includeFigCond[width=.9\linewidth]{zelefski-real-data.pdf}
  \caption[]{The results of applying the 125 combinations of learners to the
    prostate cancer data set. The learners are \texttt{KM} (KM), \texttt{Cox
      strata CS} (strata), \texttt{Lasso} (lasso), \texttt{Elastic} (elastic),
    and \texttt{RF} (RF) as described Table~\ref{tab:zel-library}. The error
    bars are based on five repetitions using different splits. We refer to
    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
    learner', `Death learner', and `Censoring learner', respectively.}
  \label{fig:zelefski-real}
\end{figure}

@@

\begin{figure}
  \centering%
  \includeFigCond[width=.8\linewidth]{zelefsky-data-target-par.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

@@

\begin{appendices}

@@

\end{appendices}

\bibliographystyle{abbrvnat}
\bibliography{bib.bib}
